Supplemental Information: Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-04-05 docs/supp_v10.Rmd
final ms edits

---
title: "Supplemental Information: Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx

@@

## Contents
This supplementary information contains 3 texts, 4 figures and 3 tables. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

@@

$\epsilon = \epsilon_S$

@@

$\epsilon = \epsilon_D$

@@

$\epsilon = \epsilon_S$

$$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
$$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$

$\epsilon = \epsilon_D$

$$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gR_hS)^{3/16}H^{3/8} \mathbf{(S16)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$

@@

$A_0_i$ and $n_i$ prior hyperparameters were assigned following @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 2.2.1 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S19)}$$

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$). However as noted in the Main Text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.

$k_{600_{\sigma}}$ is set to 0.30 (log-space). This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values. $k_{600_{\lambda}}$ was set to log(0.001) m/day. $k_{600_{\gamma}}$ was set to log(500) m/day.

Finally, recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e. stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$. For the purposes of this study, we are validating BIKER against equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$. Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30. This is the $\sigma_{k_{600}}$ used in this study.

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 9 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

@@

## Text S4: Literature $CO_2$ flux models
Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.2. The specific equations used are outlined in Table S3.

In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per equation 2. Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates. The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow. Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S3).

For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$. Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 following Text S1. We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively). Finally, we converted from $k_600$ to $k_{CO_2}$ following @raymondScalingGasTransfer2012.

@@

## Figure S2
![Figure S2 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS2.jpg)

## Figure S3
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS3.jpg)

## Figure S4
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 16 SWOT-simulated rivers with the Frasson et al. (2021) error model. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)

## Table S1
*Table S1: Studies that gas exchange velocity measurements come from. 'Study' refers to the paper from which these measurements were obtained. Any data wrangling was done by those authors. 'Workers' refers to who actually made the measurements in the field. The Raymond et al. (2012) dataset is itself a meta-analysis. Please see that paper for how those measurements were collected, and see the 'Additional Studies' for the workers who actually collected the measurements.*

@@

## Table S2
*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. Carrot accents indicate the predicted value.*

|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|
|---------------|-----------|--------------|---------------|------------------|
|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|
|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|
|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600_i}}} }{\mu_{k_{600_i}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600_i}}} / \mu _{\hat{k_{600_i}}} }{ \sigma_{k_{600_i}} / \mu_{k_{600_i}}} - 1)^2 }$|1|-∞ to 1|
|Mean Absolute Error|MAE| $\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})$ |0|0 to ∞|

## Table S3
*Table S3: Details on the 3 literature models for $CO_2$ fluxes. See Text S4 for more details.*

|**Name**|**Depth equation**|**Velocity equation**|**Description**|**Reference**|
|--------|------------------|---------------------|---------------|-------------|
|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|

>>>>

---
title: "Supplemental Information: Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt E Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
   #   pandoc_args:
  #      - '--lua-filter=scholarly-metadata.lua'
  #      - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx

@@

## Contents
This supplementary information contains 5 texts, 5 figures and 2 tables.

@@

*Log-law-of-the-wall model for the turbulent dissipation rate*

@@

*Form-drag model for the turbulent dissipation rate*

@@

*Log-law-of-the-wall model for the turbulent dissipation rate*

$$k_{600}=\beta(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
$$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$

*Form-drag model for the turbulent dissipation rate*

$$k_{600}=\beta_1(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gR_hS)^{3/16}H^{3/8} \mathbf{(S16)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$

@@

$A_0_i$ and $n_i$ prior hyperparameters were assigned following @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and statistical learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.252Q^{0.388})^{9/16}*(0.276Q^{0.164})^{1/4} \mathbf{(S19)}$$

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is the standard prior information used to benchmark the SWOT discharge algroithms and is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$). However as noted in the Main Text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.

$k_{600_{\sigma}}$ is set to 0.30 (log-space). This corresponds to a coefficient of variation of approximately 30%, which we took to reflect a reasonably strong agreement between the prior and the observed values. $k_{600_{\lambda}}$ was set to log(0.001) m/day. $k_{600_{\gamma}}$ was set to log(500) m/day.

## Text S4: Determining complete BIKER model uncertainty
Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e. stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$. For the purposes of this study, we are validating BIKER against equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$. Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30. This is the $\sigma_{k_{600}}$ used in this study.

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 9 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect, no-error measurements are made by the SWOT satellite, the full equation 7 uncertainty is expressed for some set of hydraulic observations as equation S20. To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.

More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (section 2 in the Main Text) and push 166 different Monte Carlo simulations through equation S20. Note that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates. For each of the 166 distributions of estimated $l_{600}$, we extract the standard deviation as the uncertainty. We then take the average of those standard deviations to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ and should be used when running BIKER on real SWOT data.

@@

## Text S5: Literature $CO_2$ flux models
Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.3. The specific equations used are outlined in Table S2.

In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per equation 2. Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates. The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow. Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S2).

For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$. Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering descirbed in Text S1. We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively). Finally, we converted from $k_600$ to $k_{CO_2}$ following @raymondScalingGasTransfer2012.

@@

## Figure S2
![Figure S2: Map of simulated SWOT river locations around the world. Also includes approximate location of the $CO_2$ timeseries used in this study (Figure S3). Note that river locations are approximate as some hydraulic models are not geo-referenced.](cache/maps/figS2.jpg)

## Figure S3
![Figure S3 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS3.jpg)

## Figure S4
![Figure S4: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)

## Figure S5
![Figure S5: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS5.jpg)

## Table S1
*Table S1: Studies that gas exchange velocity measurements come from. 'Study' refers to the paper from which these measurements were obtained by us. Any data wrangling was done by those authors. 'Workers' refers to who actually made the measurements in the field. The Raymond et al. (2012) dataset is itself a meta-analysis. Please see that paper for how those measurements were collected, and see the 'Additional Studies' for the workers who actually collected the measurements.*

@@

## Table S2
*Table S2: Details on the 3 literature models for $CO_2$ fluxes. See Text S5 for more details.*

|**Name**|**Depth equation**|**Velocity equation**|**Description**|**Reference**|
|--------|------------------|---------------------|---------------|-------------|
|Brinkerhoff 2019|$D=0.252Q^{0.388}$|$\bar{U}=0.276Q^{0.164}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$|@raymondScalingGasTransfer2012|1,026 measurements across the United States|@raymondScalingGasTransfer2012|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|
